Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-09-07 paper/current/main.tex
incorperate new plots into text

\begin{document} %{{{
\title{Learning a functional distance between organizations with unsupervised embeddings of scientific mobility}
\title{Unsupervised embeddings of career trajectories capture the latent structure of global scientific mobility} %{{{
\date{\today}
\maketitle %}}}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Historically, human mobility is strongly constrained by the physical separation-geographical distance~\autocite{zipf1946gravity}.
Recently, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important, which calls for alternative modeling framework for human mobility take these factors into account~\autocite{boschma2005proximity,  brown1970functional, brown1970migration, kim2018functional}.
Here, using a massive dataset of scientific mobility, we demonstrate a neural embedding model can successfully encode nuanced relationships between locations into a vector-space ~\autocite{mikolov2013word2vec}, providing an effective functional distance that reflects the multi-faced structure of scientists' mobility and capturing cultural, linguistic relations, and even academic prestige at multiple levels of granularity as well as implicit hierarchical mobility structure.
Furthermore, vector representations of organizations themselves also reveal insights into the landscape of global scientific mobility, such as a relationship between organizations' size and the norm of its vector that is consistent across most countries. 
The ability to learn scalable, dense, and meaningful high-resolution vector representations of mobility directly derived from the data can open up a new avenue of studying human mobility.

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
Although geographic distance has been constraining  human movements throughout history, it is becoming less relevant in a world connected by rapid transit and global airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a far-away country with similar language and culture, than to a much closer country such as Indonesia~\autocite{pew2018migration}.
Similarly, a student in South Korea is more likely to attend a university in Canada than one in North Korea~\autocite{unesco2019students}.
Although geographic distance has been used as the most prominent basis for models of mobility, such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} models, there have been attempts to define alternative notions of distance, or ~\textit{functional distances}~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional} from real-world data or \textit{a priori} relationships between geographic entities.
Yet, functional distances are often low-resolution, at the level of countries rather than regions, cities, or organizations, and rely on somewhat arbitrary decisions about which factors are most relevant.
Moreover, functional distances have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.
To capture such complexity of human mobility, new approaches that incorporate its many facets into a single high-resolution functional distance are needed.

Here, we propose using a neural embedding model that draws from real-world mobility trajectories can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which not only a meaningful \textit{functional distance} between places but also relationship based on culture, language, and even prestige along with geographic relationship can be probed.

@@

Here, each trajectory is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
The most fine-grained geographical unit in this data is the organization, such as a ``university'', ``research institute'', ``business'', or ``government agency''.
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).
We use the trajectories of more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.

@@

To assess the performance of the embedding distance relative to geographic distance, we evaluate their correlation with real-world mobility data and their performance in predicting real-world mobility flows as parameters to the gravity model~\autocite{zipf1946gravity, barbosa2018human}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, lewer2008immigrationgravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity} that calculates the \textit{expected flux}, $\hat{T}_{ij}$, of peoples between locations(organizations) based on their populations and distance:

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of mobile and non-mobile authors who affiliated with institution $i$, $f(r_{ij})$ is a decay function with respect to distance between organizations, and $C$ is a constant estimated from data (see Data and Methods). 
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of organizations $i$ and $j$ in the trajectory of affiliations in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$, is proportional to the organizations' sizes, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between organizations or their embedding distance.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

The embedding space also allows us to \textit{zoom in} on subsets and re-project them using UMAP to reveal local relationships.
For example, re-projecting organizations located in Western, Southern, and Southeastern Asia (Fig.~\ref{fig:projection}b) reveals a gradient of countries between Egypt and the Philippines that largely corresponds to geography, but with some exceptions along the cultural and religious spectrums;
Iraq is nearer to Malaysia in southeast-Asian than closer and more religiously-similar Saudi Arabia and Qatar.
We also validate this finding quantitatively with the distance between nations  in the embedding space $d$,  cosine distance between the centroids of organizations vectors belonging to that country. 
Then, Malaysia ($d=0.29$) is closer to Iraq more than Arabia ($d=0.50$) and Qatar ($d=0.51$).

Linguistic and historical ties also effect the scientific mobility.
We observe that Spanish-speaking post-colonial Latin American nations are positioned near Spain (Fig.~\ref{fig:projection}c), rather than nearby Portuguese-speaking nations ($d = 0.34$ vs. $d = 0.52$ for Mexico and $d = 0.39$ vs. $d = 0.49$ for Chile) reflecting linguistic and cultural ties.
Similarly, North-African countries that were once under French rule such as Morocco are closer to France ($d =0.35$) than to similarly-distant European countries such as Spain ($d = 0.40$) and Italy ($d = 0.53$).
Comparable structural patterns exist even within a single country.
For example, at the global level (Fig~\ref{fig:projection}a), organizations within Quebec in Canada are located nearer the French cluster ($d = 0.65$) than the rest of the Canadian organizations ($d = 0.88$).

@@

Just as the embedding space makes it possible to \textit{zoom in} on subsets of organizations, it is also possible to \textit{zoom out}  by aggregating organizational vectors.
We can examine the country-level structure that governs scientific mobility.
For the purpose, we define the representative vector of each country as the average of their organizational vector and perform hierarchical clustering of nations that have at least 25 organizations represented in the embedding space based on their cosine similarity, excluding the United States which is a dominant hub well connected to most other countries (see Fig. \ref{fig:clustering}a).
The six identified clusters roughly correspond to countries in East-Asia (orange), Scandinavia (dark purple), the British Commonwealth (light purple), and Central and Eastern Europe (light blue), and then two remaining diverse clusters contain a mix of European, Latin American, and Mediterranean countries.
The cluster structure shows that not only geography but also linguistic ties between countries are related to scientific mobility.

We quantify the relative importance of geography (by region), and language (by most widely-spoken language of each country) using the element-centric clustering similarity~\autocite{gates2019element}, a method that can compare hierarchical clustering by explicitly adjusting the relative importance of different levels of the hierarchy with scaling parameter $r$, acting like a \textit{zooming lens}.
If $r$ is high, the similarity is based on the lower levels of the dendrogram, whereas when $r$ is low, the similarity is based on higher levels.
Fig. \ref{fig:clustering}b demonstrates that the region plays a major role at higher levels of the clustering process (low $r$), and language (family) explains the clustering more at the lower levels (high $r$).
This suggests that the embedding space captures the hierarchical structure of mobility---at the highest level, geography defines the embedding space, followed by language.

@@

Prestige hierarchy is known to underpin the dynamics of scientific mobility, where researchers tend to move to similar or less prestigious organizations~\autocite{deville2014career, clauset2015hierarchy}. Does the embedding space, to which no prestige information is given, encode a prestige hierarchy?
We test this by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining an axis in the embedding space where poles are defined using known high- and low-ranked universities.
To define prestige of ranked U.S. ranked universities, we use the Times Ranking of World Universities (we also use research impact from the Leiden Rankings~\autocite{waltman2012leidenrankings}, see \SI);
the high-rank pole is defined as the average vector of the top five U.S. universities according to the rankings, whereas the low-ranked pole is defined using the five bottom-ranked (geographically-matched by U.S. census region) universities.
We derive an embedding-based ranking for universities based on their nearness to the high-ranked pole.

The embedding space encodes the prestige hierarchy of U.S. universities that are coherent with real-world university rankings.
The embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}a).
For reference, the embedding-based ranking achieves correlation with the Times ranking comparable to that between the Times ranking and the publication impact scores from the Leiden Ranking~\autocite{waltman2012leidenrankings}, a bibliometrically-based university ranking (Spearman's $\rho = 0.87$, Fig.~\ref{fig:semaxis}b).
The correlation between the embedding-based ranking and the Times ranking is robust regardless of the number of organizations used to define the axes (Fig. ~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities produces a ranking that is significantly correlated with the Times ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
The correlation is also comparable to network-driven measures such as degree strength (sum of edge weights, Spearman's $\rho = 0.73$) and eigenvector centrality (Spearman's $\rho = 0.76$, see \SI). 
The strongest outliers that were ranked more highly in the Times ranking than in the embedding-based ranking tend to be large state universities such as Arizona State University and  University of Florida.
Those ranked higher in the embedding based ranking tend to be relatively-small universities near major urban areas such as the University of San Francisco and University of Maryland Baltimore County.
This suggests that organizations in the embedding space are arranged according to prestige, even when the erepresentation is learned using data without explicit infomration on the direction of mobility (as in other formal models~\autocite{clauset2015hierarchy}), or prestige.

The axes can further be visualized to examine the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts to aid visualization.
Prestigious universities such as Columbia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reasoning about the prestige of organizations for which rankings are often low-resolution or incomplete, such as regional and liberal arts universities, seminaries, musical schools, and community colleges (Fig.~\ref{fig:semaxis}d), or for which no such rankings exist, such as government organizations (Fig.~\ref{fig:semaxis}d), and research institutes (Fig.~\ref{fig:semaxis}e).

@@

%
% SUBSECTION: SEMAXIS
%
\subsection*{SemAxis}
SemAxis~\autocite{an2018semaxis, nakandala2017gendered, kozlowski2018geometry} leverages the latent semantic characteristics of word embeddings to represent the position of terms along a conceptual axis, reflecting the relationship of these terms to a corresponding concept.
Each axis is an antonym pair defined for each specific use case.
In the case of natural language, the poles of the axis could be ``good'' and ``bad'', ``surprising'' and ``unsurprising'', or ``masculine'' and ``feminine''.
By treating organizations analogously to words and trajectories analogously to sentences, we can use SemAxis to leverage the semantic properties of the embedding vectors to operationalize abstract relationships between organizations.

@@

We examine the gravity model on the Personalized Page Rank (PPR)\autocite{jeh2003scaling} as a benchmark on the network. We construct the co-occurrence network of $N$ organizations, in which each edge between organizations $i$ and $j$ represents a co-occurence of $i$ and $j$ in the same affiliation trajectory, with weight $W_{ij}$ given by the sum of the co-occurences over all researchers. and edges are co-occurrence between two organizations. The Personalized Page Rank is a ranking algorithm for nodes based on a random walk process on networks. 
The walker visiting at a node moves to a neighboring node chosen randomly with a probability proportionally to the weight of the edge in one step. Furthermore, with probability $\alpha$, the walker is teleported back to the starting node. The rank of a node is determined by the probability that the walker visits the node in the stationary state. The stationary distribution of the random walker starting from node $i$, denoted by $p_i=(p_{ik})$, is given by

\begin{equation}
	\label{eq:ppr}
	p_i = (1 - \alpha) v_i + \alpha p_i W,
\end{equation}
where $v_i$ is a column vector of length $N$ with entries that are all zero except the $i$th entry that equals to one, $W = (W_{ij})$ is the weighted adjacency matrix. We used $\alpha=0.9$ here.

 We can think $p_i$ as a representation vector of the organization $i$, and calculate the distance between organizations $i$ and $j$, $d_{ij}$ with measuring distance between $p_i$ and $p_j$ to examine the gravity law. We consider two distance measures in this analysis. The first one is cosine distance which is used for our embedding method, $d_{ij} = 1 - \frac{\bm{p}_{i} \cdot \bm{p}_{j}}{\lVert \bm{p}_{i} \rVert \lVert \bm{p}_{j} \rVert}$. Also, if we think $p_i$ as a discrete probability distribution, then we can consider Jensen–Shannon divergence (JSD), can be written as,

\begin{equation}
	\label{eq:JSD}
	d_{ij} = JSD(p_i||p_j) = \frac{1}{2}D_{KL}(p_i||m) + \frac{1}{2}D_{KL}(p_j||m),
\end{equation}

\begin{equation}
	\label{eq:KL}
	D_{KL}(p_i||m) = \sum^x p_{ix}\log\frac{p_{ix}}{m_x},
\end{equation}
where $m=\frac{1}{2}(p_i+p_j)$. We report the result with cosine distance ($R^2=0.14$, Fig. \ref{fig:supp:gravity_pprcos}) and  Jensen–Shannon divergence ($R^2=0.19$, Fig. \ref{fig:supp:gravity_pprjsd}). In both cases, the performance is under the performance of the model with geographical distance. Even though the length of the PPR vectors is extremely larger than the length of our embedding vectors, result with the embedding distance outperforms both of them.

@@

>>>>

\begin{document} %{{{
\title{Learning a functional distance between organizations with unsupervised embeddings of scientific mobility}
\title{Unsupervised embeddings of organizations capture the latent structure of global scientific mobility} %{{{
\date{\today}
\maketitle %}}}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Geographic distance and population have been considered as leading factors of human mobilit~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important, which calls for alternative \textit{functional distances} that take these factors into account~\autocite{boschma2005proximity,  brown1970functional, brown1970migration, kim2018functional}.
However, existing functional distance are often arbitrary and consider only single isolated facets of mobility.
Using the case study of scientific mobility, we demonstrate the a neural embedding model can successfully encode the affiliation trajectories of scientists into a vector-space representation of organizations~\autocite{mikolov2013word2vec} from which a functional distance can be derived.
We show that the proposed functional distance reflects the multi-faced structure of scientists' mobility that is not explained by geographic distance alone, capturing relations based on culture, language, and academic prestige at multiple levels of granularity.
The vector representations of organizations themselves also reveal insights into the landscape of global scientific mobility, such as a relationship between organizations' size and the norm of its vector that is consistent across most countries. 
The ability to learn scalable, dense, and meaningful high-resolution vector representations of mobility directly from the data can open up a new avenue of studying human mobility.

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
Geographic distance has underpinned human movements, however it is becoming less relevant in a world connected by rapid transit and global airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a far-away country with similar language and culture, than to a much closer country such as Indonesia~\autocite{pew2018migration}.
Similarly, a student in South Korea is more likely to attend a university in Canada than one in North Korea~\autocite{unesco2019students}.
Although geographic distance has been used as the most prominent basis for models of mobility, such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} models, there have been attempts to define alternative notions of distance, or ~\textit{functional distances}~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional} from real-world data or \textit{a priori} relationships between geographic entities.
Yet, functional distances are often low-resolution, at the level of countries rather than regions, cities, or organizations, and rely on somewhat arbitrary decisions about which factors are most relevant.
Moreover, functional distances have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.
To capture such complexity of human mobility, new approaches that incorporate its many facets into a single high-resolution functional distance are needed.

Here, we propose using a neural embedding model that draws from real-world mobility trajectories to encode the complex landscape of global mobility into a dense, continuous, and meaningful vector-space representation, from which a \textit{functional distance} between places is derived.
Using the case of scientific mobility, we demonstrate that the functional distance captures relations based on culture, language, and academic prestige, along with geographic relations, at multiple levels of granularity

@@

Here, scientific mobility is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
The most fine-grained geographical unit in this data is the organization, such as a ``university'', ``research institute'', ``business'', or ``government agency''.
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).
We use the trajectories of more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.

@@

To assess the performance of the embedding distance relative to geographic distance, we evaluate their correlation with real-world mobility data and their performance in predicting real-world mobility flows as parameters to the gravity model~\autocite{zipf1946gravity, barbosa2018human}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, lewer2008immigrationgravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity} that calculates the \textit{expected flux}, $\hat{T}_{ij}$, of scientists between organizations based on their populations and distance:

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of mobile and non-mobile authors who affiliated with institution $i$, $f(r_{ij})$ is a decay function with respect to distance between organizations, and $C$ is a constant estimated from data (see Data and Methods). 
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of organizations $i$ and $j$ in the trajectory of affiliations in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$, is proportional to the organizations' sizes, $\hat{T}_{ij} \propto m_{i}, m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between organizations or their embedding distance.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

@@

The embedding space also allows us to \textit{zoom in} on subsets and re-project them using UMAP to reveal local relationships.
For example, re-projecting organizations located in Western, Southern, and Southeastern Asia (Fig.~\ref{fig:projection}b) reveals a gradient of countries between Egypt and the Philippines that largely corresponds to geography, but with some exceptions along the cultural and religious spectrums;
Iraq is nearer to Malaysia in southeast-Asian than closer and more religiously-similar Saudi Arabia and Qatar.
In fact, we compute the distance, $d$, between nations in the original embedding space as the cosine distance between the centroids of organizations vectors belonging to that country. 
Then, Malaysia ($d=0.29$) is closer to Iraq more than Arabia ($d=0.50$) and Qatar ($d=0.51$).
Linguistic and historical ties are also conspicuous.
We observe that Spanish-speaking post-colonial Latin American nations are positioned near Spain (Fig.~\ref{fig:projection}c), rather than nearby Portuguese-speaking nations ($d = 0.34$ vs. $d = 0.52$ for Mexico and $d = 0.39$ vs. $d = 0.49$ for Chile) reflecting linguistic and cultural ties.
Similarly, North-African countries that were once under French rule such as Morocco are closer to France ($d =0.35$) than to similarly-distant European countries such as Spain ($d = 0.40$) and Italy ($d = 0.53$).
Comparable structural patterns exist even within a single country.
For example, at the global level (Fig~\ref{fig:projection}a), organizations within Quebec in Canada are located nearer the French cluster ($d = 0.65$) than the rest of the Canadian organizations ($d = 0.88$).

@@

Just as the embedding space makes it possible to ``zoom in'' on subsets of organizations, it is also possible to ``zoom out'' by aggregating organizational vectors.
We can examine the country-level structure that governs scientific mobility.
To this end, we define the representative vector of each country as the average of their organizational vectors.
We then perform hierarchical clustering of nations that have at least 25 organizations represented in the embedding space based on their cosine similarity, excluding the United States which is a dominant hub well connected to most other countries (see Fig. \ref{fig:clustering}a).
The six identified clusters roughly correspond to countries in East-Asia (orange), Scandinavia (dark purple), the British Commonwealth (light purple), and Central and Eastern Europe (light blue), and then two remaining diverse clusters contain a mix of European, Latin American, and Mediterranean countries.
The cluster structure shows that not only geography but also linguistic ties between countries are related to scientific mobility.

We quantify the relative importance of geography (by region), and language (by most widely-spoken language of each country) using the element-centric clustering similarity~\autocite{gates2019element}, a method that can compare hierarchical clustering by explicitly adjusting the relative importance of different levels of the hierarchy with scaling parameter $r$, acting like a ``zooming lens''.
If $r$ is high, the similarity is based on the lower levels of the dendrogram, whereas when $r$ is low, the similarity is based on higher levels.
Fig. \ref{fig:clustering}b demonstrates that the region plays a major role at higher levels of the clustering process (low $r$), and language (family) explains the clustering more at the lower levels (high $r$).
This suggests that the embedding space captures the hierarchical structure of mobility---at the highest level, geography defines the embedding space, followed by language.

@@

Prestige hierarchy is known to underpin the dynamics of scientific mobility, where researchers tend to move to similar or less prestigious organizations~\autocite{deville2014career, clauset2015hierarchy}. Does the embedding space, to which no prestige information is given, encode a prestige hierarchy?
We test this by exploiting the geometric properties of the embedding space with SemAxis~\autocite{an2018semaxis}.
Here, we use SemAxis to operationalize the abstract notion of academic prestige, defining an axis in the embedding space where poles are defined using known high- and low-ranked universities.
To define prestige of ranked U.S. ranked universities, we use the Times Ranking of World Universities (we also use research impact from the Leiden Rankings~\autocite{waltman2012leidenrankings}, see \SI);
the high-rank pole is defined as the mean vector of the top five U.S. universities according to the rankings, whereas the low-ranked pole is defined using the five bottom-ranked (geographically-matched by U.S. census region) universities.
We derive an embedding-based ranking for universities based on their nearness to the high-ranked pole.

The embedding space encodes the prestige hierarchy of U.S. universities that are coherent with real-world university rankings.
The embedding-based ranking is strongly correlated with the Times ranking (Spearman's $\rho = 0.73$, Fig.~\ref{fig:semaxis}a).
For reference, the embedding-based ranking achieves correlation with the Times ranking comparable to that between the Times ranking and the publication impact scores from the Leiden Ranking~\autocite{waltman2012leidenrankings}, a bibliometrically-based university ranking (Spearman's $\rho = 0.87$, Fig.~\ref{fig:semaxis}b).
The correlation between the embedding-based ranking and the Times ranking is robust regardless of the number of organizations used to define the axes (Fig. SI~\ref{fig:supp:semaxis_compare}), such that even using only the single top-ranked and bottom-ranked universities produces a ranking that is significantly correlated with the Times ranking (Spearman's $\rho = 0.46$, Fig.~\ref{fig:supp:semaxis_compare}a).
The correlation is also comparable to network-driven measures such as degree strength (sum of edge weights, Spearman's $\rho = 0.73$) and eigenvector centrality (Spearman's $\rho = 0.76$, see \SI). 
The strongest outliers that were ranked more highly in the Times ranking than in the embedding-based ranking tend to be large state universities such as Arizona State University and  University of Florida.
Those ranked higher in the embedding based ranking tend to be relatively-small universities near major urban areas such as the University of San Francisco and University of Maryland Baltimore County.
This suggests that organizations in the embedding space are arranged according to prestige, even when the erepresentation is learned using data without explicit infomration on the direction of mobility (as in other formal models~\autocite{clauset2015hierarchy}), or prestige.

The axes can further be visualized to examine the relative position of organizations along the prestige axis, and along a geographic axis between California and Massachusetts to aid visualization.
Prestigious universities such as Columbia, Stanford, MIT, Harvard, and Rockefeller are positioned towards the top of the axis (Fig.~\ref{fig:semaxis}c).
Universities at the bottom of this axis tend to be regional universities with lower national profiles (yet still ranked by Times Higher Education) and with more emphasis on teaching, such as Barry University and California State University at Long Beach.
By projecting other types of organizations onto the prestige axis, SemAxis offers a new means of reasoning about the prestige of organizations for which rankings are often low-resolution, incomplete, or entirely absent such as regional and liberal arts universities, seminaries, musical schools, and community colleges (Fig.~\ref{fig:semaxis}d), government organizations (Fig.~\ref{fig:semaxis}d), and research institutes (Fig.~\ref{fig:semaxis}e).
Using SemAxis, we can even construct quantitative rankings for these types of organizations that correlate with their citation impact (Fig.~\ref{fig:supp:sector_semaxis_impact}.

@@

%
% SUBSECTION: SEMAXIS
%
\subsection*{SemAxis}
SemAxis~\autocite{an2018semaxis, nakandala2017gendered, kozlowski2018geometry} leverages the latent semantic characteristics of word embeddings to represent the position of terms along a conceptual axis, reflecting the relationship of these terms to a corresponding concept.
Each axis is an antonym pair defined for each specific use case.
In the case of natural language, the poles of the axis could be ``good'' and ``bad'', ``surprising'' and ``unsurprising'', or ``masculine'' and ``feminine''.
By treating organizations analogously to words and trajectories analogously to sentences, we can use SemAxis to leverage the semantic properties of the resulting embedding vectors to operationalize abstract relationships between organizations.

@@

We also examine the gravity model on the Personalized Page Rank (PPR)\autocite{jeh2003scaling} as a benchmark on the network. We construct the co-occurrence network whose nodes are organization and edges are co-occurrence between two organizations. We use 0.9 in here.

\begin{equation}
	\label{eq:ppr}
	p_i = (1.0 - \alpha) v_i + \alpha p_i W,
\end{equation}

where $v_i$ is a vector with 1 in the ith entry, whose length is the N, the number of nodes in the network, W is the weighted adjacency matrix, and $\alpha$ is the hyper-parameter is called the teleportation parameter that controls the probability that random walker revisits the original node i. We can think of this PPR vector as a representation of a node, which is N-length embedding. For examining the gravity law, we use two distance measures for PPR. The first one is cosine distance which is used for our embedding method, and the second one is the Jensen–Shannon divergence (JSD).

\begin{equation}
	\label{eq:JSD}
	JSD(P||Q) = \frac{1}{2}D_{KL}(P||M) + \frac{1}{2}D_{KL}(Q||M),
\end{equation}

\begin{equation}
	\label{eq:KL}
	D_{KL} = \sum p(x)\log\frac{p(x)}{q(x)}
\end{equation}

where $M=\frac{1}{2}(P+Q)$. We report the both results, PPR with cosine distance (Fig. \ref{fig:supp:gravity_pprcos}) and (Fig. \ref{fig:supp:gravity_pprjsd}). In both cases, the performance is under the performance of the model with geographical distance. Even though the length of the PPR vectors is extremely larger than the length of our embedding vectors, our result on the embedding outperforms than network approach. Moreover, it is also computationally efficient and salable for large datasets.

@@

%
% Figure - SemAxis By State
%
\begin{figure}[hp!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/SemAxis/sector_rank_comparison.pdf}
	\caption{
		\textbf{SemAxis reconstructs publication impact in non-university sectors.}
		Comparison between the ranking of organizations in each non-university sector by their citation impact and the embedding rank.
		Citation impact is calculated as the mean-normalized citation score using papers published in the Web of Science database between 2008 and 2019. 
		The embedding rank is derived by first projecting non-university organizations onto the SemAxis axis formed with poles defined using the top five to geographically-matched bottom five universities ranked by the 2018 Times Higher Education ranking of U.S. Universities.
		\textbf{a} Shows how the correlation between the citation impact and SemAxis rankings differ while varying the size threshold for including an organization.
		Size is calculated as the mean annualized number of unique authors publishing with that organization. 
		Annotations show the number of organizations remaining at thresholds pf 0, 50, and 100. 
		\textbf{b}. Comparison of organizations using a size threshold of 10 for regional and liberal arts colleges, and 50 for research institutes and government organizations;
		these thresholds were chosen as points thresholds of stability in \textbf{a}. 		
		The impact rank is correlated with the embedding rank for regional and liberal arts colleges with Spearman's $\rho = 0.49 (n = 48)$, research institutes with Spearman's $\rho = 0.58 (n = 159)$, and for government organizations with Spearman's $\rho = 0.36 (n = 55)$. 
	}
	\label{fig:supp:sector_semaxis_impact}
\end{figure}
